\input{header}

\thesislang{en}
\thesistitle{Algorithmic Analysis \\of Code-Breaking Games}
\thesissubtitle{Master's Thesis}
\thesisstudent{Miroslav Klimoš}
\thesiswoman{false}
\thesisfaculty{fi}
\thesisyear{2014}
\thesisadvisor{prof. RNDr. Antonín Kučera, Ph.D.}

\hypersetup{
    unicode=true,
    pdftitle={Algorithmic Analysis of Code-Breaking Games},
    pdfauthor={Miroslav Klimoš},
    %pdfkeywords={Markov decision process}{optimal scheduler}{minimal scheduler},
    colorlinks=true,        % false: boxed links; true: colored links
    linkcolor=BrickRed, 	  % color of internal links
    citecolor=BrickRed,   % color of links to bibliography
    filecolor=BrickRed, %magenta,      % color of file links
    urlcolor=BrickRed %cyan           % color of external links
}

\addbibresource{biblio.bib}

\begin{document}
\input{macros}


\FrontMatter
\setlength{\parindent}{0pt}
\ThesisTitlePage


\begin{ThesisDeclaration}
I declare that this thesis is my own work and has not been submitted
in any form for another degree or diploma at any university or
other institution of tertiary education. Information derived from the published
or unpublished work of others has been acknowledged in the text
and a list of references is given.

\AdvisorName
\end{ThesisDeclaration}


% \begin{ThesisThanks}
% I would like to express my deepest gratitude to RNDr. Tomáš Brázdil Ph.D.
%   for being my advisor,
%   for suggesting a lot of related open problems and, in particular,
%   for his understanding when I failed to solve them.
% I would also like to thank RNDr. Jan Krčál for all the discussions
%   about various problems regarding continuous-time systems.
% Last but not least I thank my family for their support through my life
%   and my friends for such a great study environment.
% \end{ThesisThanks}


\begin{ThesisKeyWords}
code braking games, \\
deductive games,\\
strategy synthesis, \\
greedy strategy,\\
SAT solving,\\
model counting, \\
mastermind, \\
counterfeit coin\\
\end{ThesisKeyWords}


\begin{ThesisAbstract}
Code-breaking games are two-player games in which the first player selects
  a code from a given set and the second player strives to reveal it using a
  minimal number of experiments.
A prominent example of a code-breaking game is the board game Mastermind,
  where the code-breaker tries to guess a combination of coloured pegs.

What strategy for experiment selection should the code-breaker use?
Is it possible to compute a lower or upper bound for the number of experiments needed?
Can we compute the optimal strategy?
What is the performance of a given heuristic?
Much research on the topic has been done but it usually focuses
  on one particular game and little has been written about code-breaking
  games in general.

In this work, we create a general model of code-breaking games
  based on propositional logic
  and design a computer language for game specification.
Most importantly, we develop a computer program for game analysis
  and strategy synthesis.
Using the tool, we can easily reproduce known results for Mastermind or
  any other code-breaking game and quickly evaluate new ideas
  for heuristics.
\end{ThesisAbstract}

\MainMatter
\setlength{\parindent}{0pt}

\setcounter{secnumdepth}{1}
\setcounter{tocdepth}{2}
\begin{spacing}{1.2} \normalsize
\tableofcontents
\end{spacing}


\input{intro}
\input{problems}
\input{formalne}
\input{cobra}
\input{implementation}
\input{experimental}
\input{conclusions}

\pagestyle{plain}
%\printbibheading
\printbibliography
\clearpage
\addcontentsline{toc}{chapter}{Bibliography}

\input{attachment}
%\clearpage
\addcontentsline{toc}{chapter}{Contents of the electronic attachment}

% \begin{appendices}
% \input{app-games}
% \end{appendices}

% \printbibliography[keyword=cc,heading=subbibliography,title={Counterfeit Coin}]
% \printbibliography[keyword=mm,heading=subbibliography,title={Mastermind}]
% \printbibliography[keyword=other,heading=subbibliography,title={Other}]
\end{document}
